---
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "pdfs") })

output:
  pdf_document:
    highlight: arrow
urlcolor: blue
linkcolor: black
---

```{r "2_expectation_variance-0", ECHO = FALSE, warning = FALSE, include = FALSE}
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(UsingR)
library(ggplot2)
```

# Expectation and variance

## Expected value and variance of common distributions
### Geometric distribution
Let $X$ be geometrically distributed with the parameter $p$. The expectation is then  
$$
E[X] = \frac{1}{p}
$$
where $p$ is the probability of success.

Let $X$ be geometrically distributed with the parameter $p$. The variance is then  
$$
Var[X] = \frac{1-p}{p^2}
$$
where $p$ is the probability of success.

### Exponential distribution
Let $X$ be exponentially distributed with the parameter $\lambda$. The expectation is then  
$$
E[X] = \frac{1}{\lambda}
$$

Let $X$ be exponentially distributed with the parameter $\lambda$. The variance is then  
$$
Var[X] = \frac{1}{\lambda^2}
$$

### Normal distribution
Let $X$ be a normally distributed random variable, or $X \sim N(\mu, \sigma2)$. The expectation is then
$$
E[X] = \mu
$$

Let $X$ be a normally distributed random variable, or $X \sim N(\mu, \sigma2)$. The variance is then
$$
Var[X] = \sigma^2
$$

## Expected value
To find the expected value of a probability distribution, we can use the following formula: 
$$
\mu \text{ or } E[X] = \sum_{i}a_i \cdot P(X = a_i) = \sum_{i} a_i \cdot p(a_i)
$$
where $a_i$ is the data value and $P(a_i)$ is the probability of that value.

**Example:**  
Here we have a sample data set of the goal probability of a football team:
```{r "2_expectation_variance-1", echo = FALSE}
library(knitr)
goals <- c(0, 1, 2, 3, 4)
probs <- c(0.18, 0.34, 0.35, 0.11, 0.02)
data  <- do.call(rbind, Map(data.frame, "Goals" = goals, "Probs" = probs))

colnames(data) <- c("Goals (X)", "Probability P(X)")

kable(data,
  caption = "Goal probability of a football team", booktabs = T) %>%
  kable_styling(latex_options = "HOLD_position")
```

### Method 1 - using `weighted.mean()`
The first method to calculate the expected value is using the `weighted.mean()` function as shown:

```{r "2_expectation_variance-2"}
#define values
vals <- c(0, 1, 2, 3, 4)

#define probabilities
probs <- c(.18, .34, .35, .11, .02)

#calculate expected value
weighted.mean(vals, probs)
```

### Sample mean
We can calculate the sample mean using the `mean()` function like this:

```{r "2_expectation_variance-3"}
#define values
vals <- c(3, -2, -5, 2, 5, 2, 5, -1, -3, 4, 2)

#calculate sample mean
mean(vals)
```

## Variance
The variance is a way to measure how spread out data values are around the mean.
The formula to find the variance of a population is as follows:
$$
\sigma^2 \text{ or } Var(X) = E[X^2] - E[X]^2
$$
If the generator of the random variable X is discrete with a probability mass function ($x_1 \rightarrow p_1, x_2 -> p_2 ... x_n -> p_n$) then the variance can be calculated as such:
$$
Var(X) = \sum_{i=1}^n p_i \cdot (x_i - \mu)^2
$$
Where $\mu$ is the expected value i.e.:
$$
\mu \text{ or } E[X] = \sum_{i}a_i \cdot P(X = a_i) = \sum_{i} a_i \cdot p(a_i)
$$
where $a_i$ is the data value and $P(a_i)$ is the probability of that value.

### Sample variance

Suppose we have the following dataset:

```{r "2_expectation_variance-4", echo = FALSE}
data <- c(2, 4, 4, 7, 8, 12, 14, 15, 19, 22)
kable(t(data),
  caption = "Sample dataset", booktabs = T) %>%
  kable_styling(latex_options = "HOLD_position")
```

We can then calculate the sample variance using the `var()` function:

```{r "2_expectation_variance-5"}
#define dataset
data <- c(2, 4, 4, 7, 8, 12, 14, 15, 19, 22)

#calculate sample variance
var(data)
```

### Population variance

Suppose we have the same dataset as previously we can then calculate the population variance by simply multiplying the sample variance by `(n-1)/n`:

```{r "2_expectation_variance-6"}
data <- c(2, 4, 4, 7, 8, 12, 14, 15, 19, 22)

#determine length of data
n <- length(data)

#calculate population variance
var(data) * (n - 1) / n
```