---
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "pdfs") })

output:
  pdf_document:
    highlight: arrow
urlcolor: blue
---

```{r "4_covariance_correlation-0", ECHO = FALSE, warning = FALSE, include = FALSE}
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(UsingR)
library(ggplot2)
```

# Covariance and correlation

## Covariance
Covariance is a measure of how changes in one variable are associated with changes in a second variable. Specifically, it's a measure of the degree to which two variables are linearly associated.

A covariance matrix is a square matrix that shows the covariance between many different variables. This can be a useful way to understand how different variables are related in a dataset.

### Covariance between two variables

```{r "4_covariance_correlation-1"}
data <- data.frame(math = c(84, 82, 81, 89, 73, 94, 92, 70, 88, 95),
                   science = c(85, 82, 72, 77, 75, 89, 95, 84, 77, 94),
                   history = c(97, 94, 93, 95, 88, 82, 78, 84, 69, 78))

# create the covariance matrix for the two varibles "math" and "science"
cov(data[, c("math", "science")])
```
Explanation of the covariance matrix in the following section.

### Covariance between multiple variables
```{r "4_covariance_correlation-2"}
# First, we’ll create a data frame that contains the test scores of 10 different
# students for three subjects: math, science, and history.
data <- data.frame(math = c(84, 82, 81, 89, 73, 94, 92, 70, 88, 95),
                   science = c(85, 82, 72, 77, 75, 89, 95, 84, 77, 94),
                   history = c(97, 94, 93, 95, 88, 82, 78, 84, 69, 78))

# create the covariance matrix
cov(data)
```

The numbers on the diagonal of the covariance matrix are the variances of the different variables in the data frame. Meaning that:

* The math scores have a variance of $72.177$
* The science scores have a variance of $62.666$
* The history scores have a variance of $83.955$

The other values in the matrix represent the covariances between the various subjects. For example:

* The covariance between the math and science scores is $36.89$
* The covariance between the math and history scores is $-27.16$
* The covariance between the science and history scores is $-26.78$

A positive number for covariance indicates that two variables tend to increase or decrease in tandem. For example, math and science have a positive covariance ($36.89$), which indicates that students who score high on math also tend to score high on science. Conversely, students who score low on math also tend to score low on science.

A negative number for covariance indicates that as one variable increases, a second variable tends to decrease. For example, math and history have a negative covariance ($-27.16$), which indicates that students who score high on math tend to score low on history. Conversely, students who score low on math tend to score high on history.

```{r "4_covariance_correlation-3"}
# if you want to make sure that you only calculate the correlation numerical
# variables you can use this function:
cor(data[, unlist(lapply(data, is.numeric))])
```

## Correlation

One way to quantify the relationship between two variables is to use the Pearson correlation coefficient, which is a measure of the linear association between two variables. It always takes on a value between $-1$ and $1$ where:

$-1$ indicates a perfectly negative linear correlation between two variables
$0$ indicates no linear correlation between two variables
$1$ indicates a perfectly positive linear correlation between two variables

### Correlation between two variables

To calculate the correlation between two variables you can use the `cor(x, y)` function.

```{r "4_covariance_correlation-5"}
#create data frame
df <- data.frame(A = c(2, 3, 3, 5, 6, 9, 14, 15, 19, 21, 22, 23),
                 B = c(23, 24, 24, 23, 17, 28, 38, 34, 35, 39, 41, 43),
                 C = c(13, 14, 14, 14, 15, 17, 18, 19, 22, 20, 24, 26),
                 D = c(6, 6, 7, 8, 8, 8, 7, 6, 5, 3, 3, 2))
cor(df$A, df$B)
```
This means that there is a strong positive linear correlation between the two variables, because it is almost $1$ (1 means that there is a perfect positive linear correlation between two variables)
$`r round(cor(df$A, df$B), 2)`$ 

### Correlation between multiple variables

You can calculate the correlation between a subset of the variables in your dataset e.g.:

```{r "4_covariance_correlation-6"}
cor(df[c("A", "B", "D")])
```

The way to interpret the output is as follows:

* The correlation between A and B is  $0.9279869$.
* The correlation between A and D is $-0.7915488$.
* The correlation between B and D is $-0.7917973$.

Meaning that A and D have a somewhat strong negative linear correlation. B and D also have some strong negative linear correlation and lastly A and B have a strong positive linear correlation.

You can also calculate the correlation all of the variables in your dataset e.g.:

```{r "4_covariance_correlation-7"}
cor(df)

# if you want to make sure that you only calculate the correlation numerical
# variables you can use this function:
cor(df[, unlist(lapply(df, is.numeric))])
```